---
title: 'Graphical representation example: Alternative interpretations'
output:
  html_document: 
    df_print:        paged
    code_folding:    hide
    toc:             yes
    toc_float:       yes
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(knitr)
library(patchwork)
library(rstudioapi)

ROOT_DIR <- getActiveProject()
DOC_DIR  <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      =   300,
  dev.args = list(type = "cairo-png")
)

# Output configuration options:
options(digits = 3)
```

# Introduction

After the 2023/11/02 discussion, there are two issues that remain unclear
regarding the graphical representation of the items.

1. Whether the correlation-based or the covariance-based matrix should be used
in the graphical representation, and whether it has an effect on its
interpretation.

2. The need to represent the items in the latent vs. the test space.

We have a look at the example changing these two issues to see the effect.

# Preface

```{r read-chunks-graphical-example}
#| echo: false
read_chunk("src/Graphical_example_paper.R")
```

Load libaries:

```{r libraries}
```

Source R scripts:

```{r sources}
```

Define constants:

```{r constants}
```

Configure graphical output:

```{r graphical-output-conf}
```

# Covariance-based parameters and representation

As I say in the draft, my understanding is that the correlation-based version of
the parameters should be used for the graphical representation, as these are
scale preserving, and therefore can be interpreted straightforwardly.

We first compute the original item parameters and coordinates:

```{r compute-example-items}
```

## MCLM parameters

For reference, the MCLM version of the parameters, along with the "orthogonal"
(i.e. agnostic) ones, are:

```{r example-items-table-out}
item_params |> select(-ends_with("_obl"))
```

## Parameters and representation in the original oblique space

The graphical representation (only the oblique version is shown), as computed in
the draft, is:

```{r compose-oblique-plot}
```

```{r compose-orthogonal-plot}
```

```{r item-plot-out}
#| fig.cap: |
#|   Item plots in an oblique test space, with original (i.e., standard)
#|   covariance matrix
plot_oblique + scale_color_manual(values = PALETTE)
```

And the correlation-based parameters and coordinates:

```{r example-items-obl-out}
items_oblique
```

## Parameters and representation in oblique space with different variances

Let's now compute a covariance-based version of the parameters. In this case,
we defined the variances of the two latent dimensions as 1, so the
covariance-based and the correlation-based versions of the indices are the same.
In order to solve this, we define a new covariance matrix where the variance are
different from 1. We set them to $[2, 1.5]$ instead:

```{r define-cov-matrix}
VARIANCES  <- c(2, 1.5)
sd_matrix  <- VARIANCES |> sqrt() |> diag()
cov_matrix <- sd_matrix %*% corr_matrix %*% sd_matrix
```

With this covariance matrix, we compute the item parameters again:

```{r recompute-oblique-items}
## Compute parameters and coordinates:
items_oblique_params <- items_M2PL |> compute_mirt_params(
  all_of(INTERCEPT_COLKEY), starts_with(DISCR_PREFFIX), # Input parameters
  cov_matrix = cov_matrix,                              # Input covariances
  dir_out    = c(COSINE_DIRTYPE, DEGREE_DIRTYPE)        # Output configuration
)
items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
  D, MDISC, starts_with(COSINE_DIRTYPE),
  transform       = transform_matrix,
  original_coords = FALSE
)
items_oblique        <- full_join(
  items_oblique_params, items_oblique_coords,
  by = ITEM_COLKEY
)

# Create the parameter table again:
item_params <- items_M2PL |> full_join(
  items_oblique |> select(item:deg_2, -starts_with(COSINE_DIRTYPE)),
  by = ITEM_COLKEY
)
```

```{r example-items-obl-out}
```

As can be seen, the directions change, due to the difference in the scale of the
two dimensions induced by the different variances. The resulting graphical
representation is:

```{r compose-oblique-plot}
```

```{r item-plot-cov-out}
#| fig.cap: |
#|   Item plots in an oblique test space, with variances [2, 1.5]
plot_oblique +
  scale_x_continuous(
    limits       = c(-2, 4.2),
    breaks       = (0:4) - 2 / CORR_ARC_TAN,
    labels       = 0:4,
    minor_breaks = NULL,
    name         = NULL,
    oob          = oob_keep
  )          +
  scale_color_manual(values = PALETTE)
```

### Interpretation

The vectors are "stretched" in both directions, due to the different variances.
The "length" by which they are scaled depends on the standard deviation of each
dimension. In this case, we are using variances of 2 and 1.5. Therefore, the
vectors are stretched more in the first (horizontal) direction than in the
second (oblique) one.

## Case with more predictable results

My expectations were that the representation if the items would be independent
of the covariance matrix, but this has not been the case. I had not predicted
that the item vectors would be "stretched" in space, as has happened, although I
should have expected this because I mention it in the draft, in section 3.3
(Properties of the parameters). Therefore, I acknowledge this representation is
the correct one.

In order to test the prediction that the items are scaled by the standard
deviation, let's test a case that is easier to predict. In order to it, I will
use variances of 0.25, which result in standard deviations of 0.5. Then, if my
understanding of the representation of the vectors is correct, all the vectors
should be scaled by $tfrac{1}{2}$; that is, they should have half the original
length. Their direction however should not change.

```{r define-cov-matrix-prediction}
VARIANCES  <- c(.25, .25)
sd_matrix  <- VARIANCES |> sqrt() |> diag()
cov_matrix <- sd_matrix %*% corr_matrix %*% sd_matrix
```

```{r recompute-oblique-items}
```

```{r example-items-obl-out}
```

```{r compose-oblique-plot}
```

```{r item-plot-cov-prediction-out}
#| fig.cap: |
#|   Item plots in an oblique test space, with variances [0.25, 0.25]
plot_oblique + scale_color_manual(values = PALETTE)
```

### Interpretation

The resulting MIDSC parameters are effectively half the length, as predicted.
On the other hand, as the MIL is affected by the inverse of the MDISC, the MIL
values are the double of the original ones. This result in vectors that are not
only half the length of the originals; their origin is twice as far from
the origin of the latent space.

## Conclusions

The graphical representation reflects correctly the values of the parameters.
However, the coordinates of the parameters in the latent space do not reflect
the values of the original MCLM parameters.

This is the reason why I understand that a correlation-based representation is
more appropriate.

## Using the "covariance-based" transformation matrix

After pondering the results in the previous sections, I have realized the
following: Although the parameters and the coordinates in the original space are
correctly computed, when I transform the coordinates into the rectangular
coordinate system the transformation (i.e. mapping?) matrix I am using is not a
square root of the inverse covariance matrix, as I state in the algorithm for
computing the graphical representation (section 3.4 in the draft). Instead, what
I was using was the square root fo the inverse correlation matrix, therefore
applying a transformation that intends to be "scale invariant". As a result, the
graphical representation shows the parameters scaled by the standard deviation
of each dimension.

However, I foresee that the following will happen: If I use the _correct_ square
root matrix, that is, the square root matrix of the inverse of the covariance
matrix, the graphical representation will be always the same, independently of
the variances of the dimensions (it will only depend on the correlations).

Let's try with the previous two examples, computing first the square root
matrix. Given that $\mathbf{\Sigma} = \mathbf{SRS}$:

$$\mathbf{\Sigma}^{-1} = (\mathbf{SRS})^{-1};$$

$$\mathbf{\Sigma}^{-1} = \mathbf{S}^{-1}\mathbf{R}^{-1}\mathbf{S}^{-1};$$

$$\mathbf{P^TP} = \mathbf{S}^{-1}\mathbf{P^T_RP_R}\mathbf{S}^{-1},$$

where $\mathbf{P_R}$ refers to the (previously used) correlation-based
transformation matrix. Therefore, we can equate[^mat-eq]
$\mathbf{P} = \mathbf{P_R}\mathbf{S}^{-1}$.

[^mat-eq]: They are not necessarily the same, but we can define them to be for
our convenience, as defining it that way allegedly keeps the desired property of
leaving the horizontal axis unrotated.

With this formula, we compute the new (covariance-based) transformation matrix:

```{r recompute-transform-matrix}
cb_transform_matrix <- transform_matrix %*% solve(sd_matrix)
```

Let's verify that this matrix is a square root of the inverse covariance matrix

```{r check-transform-matrix}
t(cb_transform_matrix) %*%
  cb_transform_matrix  |>
  solve()              |>
  all.equal(cov_matrix)
```

Both matrices are the same (with double numerical precision).

Let's now try the latter example with this new transform matrix.

```{r recompute-oblique-coords}
## Compute coordinates:
items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
  D, MDISC, starts_with(COSINE_DIRTYPE),
  transform       = cb_transform_matrix,
  original_coords = FALSE
)
items_oblique        <- full_join(
  items_oblique_params, items_oblique_coords,
  by = ITEM_COLKEY
)

# Create the parameter table again:
item_params <- items_M2PL |> full_join(
  items_oblique |> select(item:deg_2, -starts_with(COSINE_DIRTYPE)),
  by = ITEM_COLKEY
)
```

```{r example-items-obl-out}
```

```{r compose-oblique-plot}
```

```{r item-plot-cov-based-transform-out}
#| fig.cap: |
#|   Item plots in an oblique test space, with variances [0.25, 0.25], and using
#|   the "covariance-based transformation matrix.
plot_oblique + scale_color_manual(values = PALETTE)
```

### Interpretation

The discrimination parameters are "rescaled" back to their original value.
However, the shift of the location is even more exaggerated due to being
scaled _again_ by a factor dependent on the inverse of the variances. Because of
this, the new location in the graphic is thus meaningless, in terms of the
interpretation of the multidimensional parameters.

## Conclusion

There are two ways to represent the item vectors in the bidimensional space:

1. Using the correlation-based definition of the multidimensional parameters.

2. Using the covariance-based definition of the multidimensional parameters.

The first one has the advantage that the items are represented in the original
scale of the test space, assuming a distribution with the given variances; the
original MCLM parameters are therefore kept invariant. The second one implies
that _the items are normalized to a standard distribution_; _the original MCLM
parameters are thus scaled accordingly_ (see [Section 3.3][gh-section-ref] of
the current draft), and the representation of the distribution in this case
would be made with a unitary variance.

[gh-section-ref]: https://github.com/DaniMori/mirt-parameters/blob/8fa41540ff077f07db4b87840f0acc86eb4be2ea/output/paper_draft/paper_M2PLM_parameters.Rmd#L676

In each of these two cases though, the oblique coordinates must be transformed
back to rectangular coordinates _using a transformation matrix that is a
square-root matrix of the correlation and **not the covariance matrix**_.

# Representation in the latent space instead of the test space

It is a concern that the multidimensional parameters should be defined
(according to Reckase's, -@reckase_difficulty_1985 definition), in the latent
space. All the computations (both the ones here and in the draft) assume that
the representation of the vectors are in the test space, after arguing that this
should be the correct representation instead of the one in the latent space.

It should be argued (probably more clearly in the draft) that the fact that
Reckase proposes the representation in the latent space may be due to the
confusion resulting from considering that both constitute a single entity. This
is because in the orthogonal case they overlap, making them indistinguishable in
practice.

Let's try to represent these sample items in the latent space, in order to
explore this representation.

## Parameters and representation in the oblique space

The MDISC and the distance component of the MIL are the same in both spaces.
However, in order to represent the item in the latent space, instead of
[Equation 25][gh-eq-25], we must apply [Equation 20][gh-eq-20] to compute the
angles with the axes of the latent space. In order to that, we first need to 
redefine the function [`compute_mirt_params()`][gh-par-fun] to compute the
latent space angles instead of the test space ones:

[gh-eq-25]: https://github.com/DaniMori/mirt-parameters/blob/8fa41540ff077f07db4b87840f0acc86eb4be2ea/output/paper_draft/paper_M2PLM_parameters.Rmd#L599

[gh-eq-20]: https://github.com/DaniMori/mirt-parameters/blob/8fa41540ff077f07db4b87840f0acc86eb4be2ea/output/paper_draft/paper_M2PLM_parameters.Rmd#L532

[gh-par-fun]: https://github.com/DaniMori/mirt-parameters/blob/8fa41540ff077f07db4b87840f0acc86eb4be2ea/R/Mirt_toolbox.R#L30

```{r redefine-compute-mirt-params}
compute_mirt_params <- function(items,
                                intercept,
                                discrimination,
                                cov_matrix,
                                item_id = item,
                                dir_out = c("cos", "rad", "deg")) {
  ## Argument checking and formatting: ----
  
  # Enquote column selections:
  item_id        <- rlang::enquo(item_id)
  intercept      <- rlang::enquo(intercept)
  discrimination <- rlang::enquo(discrimination)
  
  # Check number of columns selected:
  
  id_col     <- items |> dplyr::select(!!item_id)
  intr_col   <- items |> dplyr::select(!!intercept)
  discr_cols <- items |> dplyr::select(!!discrimination)
  
  n_discr_cols <- discr_cols |> ncol()
  
  assertive.properties::assert_is_of_length(id_col,   1)
  assertive.properties::assert_is_of_length(intr_col, 1)
  
  assertive.extra::assert_is_greater_than_or_equal_to(n_discr_cols, 1)
  
  
  # Check/format covariance matrix:
  if (missing(cov_matrix)) {
    
    cov_matrix <- n_discr_cols |> diag()
    
  } else {
    
    dim_cov <- cov_matrix |> ncol()
    
    assertive.base::assert_are_identical(dim_cov, n_discr_cols)
    assertive.extra::assert_is_covariance_matrix(cov_matrix)
  }
  
  # Direction options:
  dir_out <- dir_out |> match.arg(several.ok = TRUE)
  one_dir <- dir_out                      |>
    assertive.properties::is_of_length(1) |>
    as.logical()

  ## Main: ----
  
  # Compute Gram (i.e. inner product) matrix and "scaling" (D) matrix:
  gram_matrix  <- cov_matrix |> solve()
  scale_matrix <- gram_matrix |> diag() |> diag() |> sqrt() |> solve()

  items                                                                      |>
    tidyr::pivot_longer(!!discrimination, names_to = "par", values_to = "a") |>
    dplyr::group_by(!!item_id)                                               |>
    # This renaming is necessary for accessing the variable in `transmute`:
    dplyr::rename(d = !!intercept)                                           |>
    dplyr::transmute(
      !!item_id,
      ## Compute the multidimensional parameters:
      dim   = par |> stringr::str_extract("\\d+"),
      MDISC = (t(a) %*% cov_matrix %*% a) |> drop() |> sqrt(),
      D     = - d / MDISC,
      cos   = (scale_matrix %*% a / MDISC) |> drop(),
      rad   = dplyr::if_else(
        assertive.numbers::is_equal_to(cos, 1), # If cos is a "computational 1",
        0,                                  # then fill in its arc-cos with a 0;
        suppressWarnings(acos(cos))   # otherwise, compute its arc-cos normally.
      ),
      deg   = rad * RAD_DEG_FACTOR
    )                                                                        |>
    dplyr::ungroup()                                                         |>
    select(!!item_id, dim, MDISC, D, all_of(dir_out))                        |>
    tidyr::pivot_wider(
      names_from  = dim,
      values_from = all_of(dir_out),
      names_glue  = one_dir |> if_else('{.value}_{.name}', '{.name}')
    )
}
```

In the previous function, we have inverted the roles of the covariance matrix
(i.e., the Gram matrix in the test space) and the "Gram matrix" of the latent
space, which is the inverse of the covariance matrix. Also, we have computed a
"scale matrix" that is formed by the square root elements of the diagonal of the
Gram matrix, inverted (i.e. matrix $\mathbf{D}$ in the manuscript). Note that,
contrary to what happens in the test space (where the "scale" matrix is the
inverse of the standard deviation matrix), this scale matrix in the latent space
is "meaningless" (in terms of interpretation). Finally, in the test pace, the
covariance matrix was used in the computation of the cosines, but here the Gram
matrix is not; this makes sense since, as we have seen before, transforming from
the test to the latent space implies multiplying by the inverse of the
covariance matrix.

We now compute the coordinates in the latent space. In order to compute the
transformed coordinates, instead of the transformation matrix, I understand we
need its "complementary" with the same role in the latent space. That would be
the inverse transposed of that transformation matrix.

We return in this case the original coordinates (in the latent space) along with
the transformed ones (to rectangular coordinates).

```{r compute-oblique-items-latent-space}
## Compute parameters and coordinates:
items_oblique_params <- items_M2PL |> compute_mirt_params(
  all_of(INTERCEPT_COLKEY), starts_with(DISCR_PREFFIX), # Input parameters
  cov_matrix = corr_matrix,                             # Input covariances
  dir_out    = c(COSINE_DIRTYPE, DEGREE_DIRTYPE)        # Output configuration
)
items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
  D, MDISC, starts_with(COSINE_DIRTYPE),
  transform       = transform_matrix_inv_transp,
  original_coords = TRUE
)
items_oblique        <- full_join(
  items_oblique_params, items_oblique_coords,
  by = ITEM_COLKEY
)

# Create the parameter table again:
item_params <- items_M2PL |> full_join(
  items_oblique |> select(item:deg_2, -starts_with(COSINE_DIRTYPE)),
  by = ITEM_COLKEY
)
```

```{r example-items-obl-out}
```
```{r compose-oblique-plot-latent-space}
plot_oblique <- items_oblique |>
  arrange(desc(item))         |> # To plot them in reverse order
  ggplot(
    aes(
      origin_transf_1, origin_transf_2,
      xend  = end_transf_1, yend = end_transf_2,
      color = item, fill = item
    ),
  )                                                         +
  geom_segment(
    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin  = "mitre",
    linewidth = VECTOR_WIDTH
  )                                                         +
  scale_x_continuous(
    limits       = c(-2, 3),
    breaks       = (0:4) - 2 / CORR_ARC_TAN,
    labels       = 0:4,
    minor_breaks = NULL,
    name         = NULL,
    oob          = oob_keep
  )                                                         +
  scale_y_continuous(
    limits       = c(-2, 2.8),
    breaks       = -3:3 * cos(CORR_VALUE),
    labels       = function(x) x / cos(CORR_VALUE),
    minor_breaks = 0,
    name         = NULL,
    oob          = oob_keep
  )                                                         +
  geom_abline(slope = CORR_ARC_TAN, linewidth = LINE_WIDTH) +
  coord_fixed(expand = FALSE, clip = "on")                  +
  scale_color_manual(values = PALETTE)                +
  geom_abline(
    slope     =  CORR_ARC_TAN,
    intercept = -CORR_ARC_TAN * (-4:4),
    linewidth = LINE_WIDTH,
    linetype  = "17"
  )                                                         +
  geom_hline(yintercept = 0, linewidth = LINE_WIDTH)        +
  theme(
    axis.line          = element_blank(),
    panel.grid.major.y = element_line(
      color     = "black",
      linewidth = LINE_WIDTH,
      linetype  = "17"
    )
  )
```

```{r item-plot-latent-space-out}
#| fig.cap: |
#|   Item plots in an oblique, orthonormal latent space (with correlation = .5)
plot_oblique
```
