---
title: A generalization of multidimensional item response theory parameters
author: "Daniel Morillo, Mario Luzardo"
date: "2023-02-27"
output: word_document
bibliography: ../../www/Multidimensional-parameters-MCLM.bib
csl:          ../../www/apa-old-doi-prefix.csl
---

# Abstract

```{r setup}
#| include: false

# Includes:

library(knitr)
library(rstudioapi)

ROOT_DIR <- getActiveProject()
DOC_DIR  <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  message  = FALSE,
  cache    = FALSE
)
```

```{r sources}
#| cache:   false
#| include: false
source("R/Constants.R", encoding = 'UTF-8')
```

The Multidimensional 2-Parameter Logistic
[`r MODEL_ACRONYM`, @mckinley_extension_1983] model is one of the most important
and widely used models in Multidimensional Item Response Theory (MIRT).
@reckase_difficulty_1985 and @reckase_discriminating_1991 gave a precise
definition of a Multidimensional Item Difficulty (MID) and Multidimensional
Discrimination (MDISC) parameter, respectively, and derived their expressions
for the M2PL model.
However, we argue here that (1) their derivations implicitly assume the
orthonormality of the latent space, and (2) it is straightforwardly extensible
to applications where the items have marginally unchanging monotony, instead of
just to monotonously increasing items (which is important for non-cognitive
applications).

We derive generalized expressions of those parameters, that apply to any item
with unchanging monotony in any non-orthonormal space, thus allowing to properly
represent disparate covariance structures.
We then use these definitions in two applications: to represent M2PL items in
non-orthogonal spaces, and to compute the multidimensional parameters in various
applications, showing how they provide better information compared with their
"covariance-agnostic" counterparts.

These results become increasingly relevant as the modern measurement theory in
social and behavioral sciences experiences as a convergence between MIRT and the
common factor model. Nevertheless, we warn against generalizing our results to
other MIRT models, beyond the M2PL, without carefully considering each
particular case.

# References
