---
title: IMPS 2025 proposal
author: "Daniel Morillo, Mario Luzardo"
date: "2023-02-27"
output: word_document
bibliography: ../../www/Multidimensional-parameters-MCLM.bib
csl:          ../../www/apa-old-doi-prefix.csl
---

```{r setup}
#| include: false

# Includes:

library(knitr)
library(here)

ROOT_DIR <- here()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  message  = FALSE,
  cache    = FALSE
)
```

```{r sources}
#| cache:   false
#| include: false
source("R/Constants.R", encoding = 'UTF-8')
```

# Format

## Which presentation format would you prefer?

- [X] Oral

- [ ] Poster

- [ ] Symposia

- [ ] Symposium Overview 

# Title & Abstract

## Title

A generalization of multidimensional item response theory parameters

<!-- # 8/10 words -->

## Abstract

The Multidimensional 2-Parameter Logistic (`r MODEL_ACRONYM`, McKinley &
Reckase, 1983) model is one of the most important and widely used models in
Multidimensional Item Response Theory (MIRT).
Reckase (1985) and Reckase & McKinley (1991) gave a precise definition of a
Multidimensional Item Difficulty (MID) and Multidimensional Discrimination
(MDISC) parameter, respectively, and derived their expressions for the M2PL
model.
However, we argue that their derivations (1) only consider the case of items
with conditional monotonously increasing probability of a positive response, and
(2) implicitly assume that the items are represented in an orthogonal basis.

We argue that these assumptions, which restrict the application of the model to
the assessment of non-cognitive traits, can be relaxed.
We thus extend the expressions to the general case of unchanging monotony items
in any non-orthonormal space, which allows to properly represent disparate
covariance structures.
We then use these definitions in two applications: to represent M2PL items in
non-orthonormal bases, and to compute the multidimensional parameters in various
applications, showing how they provide better properties compared with their
"covariance-agnostic" counterparts.

These results become increasingly relevant as the modern measurement theory in
social and behavioral sciences experiences as a convergence between MIRT and the
common factor model.
Importantly, our derivations show that the space where the items are represented
is necessarily distinct from the latent space.
Nevertheless, we warn against generalizing our results to
other MIRT models (i.e. beyond the M2PL), without carefully considering each
particular case.

<!-- # 239/250 words -->

# Authors


## Is this a student submission?

- [ ] Yes

- [X] No

## Please nominate all the authors that contributed to this submission below.

Daniel Morillo

danielmorillo.ac@gmail.com

### Title

Dr

### Affiliation

Universidad Nacional de Educación a Distancia (UNED)

### Country

Spain

## Add another author

### Please type the new author's email

mluzardov@gmail.com

### Title

Dr

### First Name

Mario

### Last Name

Luzardo-Verde

### Affiliation

Universidad de la República (UdelaR)

### Country

Uruguay

# Topics

Please choose a maximum of three topics from the list below.

- [ ] Applications
- [ ] Artificial Intelligence Applications
- [ ] Bayesian Statistics
- [ ] Categorical Data Analysis
- [ ] Causal Inference and Mediation
- [ ] Classical Test Theory
- [ ] Classification
- [ ] Clustering
- [ ] Cognitive Diagnostic Models
- [ ] Conceptual and Philosophical Issues
- [ ] Computational modeling
- [ ] Differential Item Functioning
- [ ] Dynamical Systems Modeling
- [ ] Computer-based Testing
- [ ] Estimation and Computational Methods
- [ ] Factor Analysis
- [ ] Generalized Linear Models
- [ ] Graphical Models
- [X] Item Response Theory
- [ ] Latent Class Models
- [ ] Longitudinal Data Analysis
- [ ] Measurement Invariance and DIF
- [ ] Missing Data
- [X] Multidimensional Analysis
- [ ] Multilevel/Hierarchical/Mixed Models
- [ ] Multivariate Analysis
- [ ] Social Network Analysis
- [ ] Prediction
- [ ] Network Psychometrics
- [ ] Optimal/Dual Scaling
- [ ] Principal Components/Correspondence Analysis
- [ ] Patient-Reported Outcomes
- [ ] Statistical and Machine Learning
- [ ] Test Equating
- [ ] Time Series Analysis
- [ ] Validity and Reliability

# Additional Information

Links to author's websites.

Please provide a link to any professional websites you would like to share in the app.

# Application for Student Travel Award

(NA)

Proceedings

- [ ] I plan to submit a chapter based on my presentation for the 2025 IMPS Proceedings volume.

<!-- # TODO: Decide -->

# References
